---
title: "Historical Review"
format: html
---

# Financial Transaction Analysis

This section presents a dynamic analysis of the club's financial activities. The underlying data is sourced directly from a real-time ledger maintained in Google Sheets. The visualizations herein are programmatically generated and will automatically update to reflect any modifications to the source ledger, ensuring continuous and accurate financial reporting.

```{r transaction-analysis-from-google-sheet, message=FALSE, warning=FALSE}
# Load necessary libraries
library(googlesheets4)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(stringr)

# --- Authenticate and Load Data from Google Sheets ---
# This allows the script to read a public sheet without requiring a login each time.
gs4_deauth()

# The URL of your Google Sheet
sheet_url <- "https://docs.google.com/spreadsheets/d/1QUmtSwyRjy1oQ6fqXGbksyNcvsLl_8wC0t6PSgTduJ8/edit?usp=sharing" 

# Read the data from the first sheet of your Google Sheet
raw_data <- read_sheet(sheet_url, sheet = 1)

# Standardize column names for easier use
colnames(raw_data) <- make.names(colnames(raw_data), unique = TRUE)

cleaned_data <- raw_data %>%
  # Ensure Date column is in Date format
  mutate(Date = as.Date(Date)) %>%
  # Rename columns for clarity. The data is already clean.
  rename(Amount_Num = Amount, Total_Num = Total) %>%
  # Add a transaction type column
  mutate(Type = ifelse(Amount_Num >= 0, "Income", "Spending")) %>%
  # Filter out any rows where key data is missing
  filter(!is.na(Date) & !is.na(Amount_Num))

# Filter out the beginning balance for the income/spending chart
transactions <- cleaned_data %>%
  filter(Vendor != "Beginning balance")

# --- Visualization 1: Balance Over Time ---
ggplot(cleaned_data, aes(x = Date, y = Total_Num)) +
  geom_line(color = "#0073C2", size = 1) +
  geom_point(color = "#0073C2", size = 3) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Club Balance Over Time (2024-2025)",
    x = "Date",
    y = "Total Balance"
  ) +
  theme_minimal(base_size = 14)

# --- Visualization 2: Income vs. Spending ---
ggplot(transactions, aes(x = Date, y = Amount_Num, fill = Type)) +
  geom_col(position = "identity") +
  scale_fill_manual(values = c("Income" = "#4CAF50", "Spending" = "#F44336")) +
  scale_y_continuous(labels = dollar_format()) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Income and Spending Transactions",
    x = "Date",
    y = "Amount"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank())

# --- Spending Pie Chart ---
# Step 1: Isolate and categorize spending data.
spending_categorized <- cleaned_data %>%
  filter(Amount_Num < 0) %>%
  mutate(
    Category = case_when(
      str_detect(Description, regex("trip|rooms|Inn|Suites|van|Hotel", ignore_case = TRUE)) ~ "Travel & Lodging",
      str_detect(Description, regex("Balls|Uniforms|Merch", ignore_case = TRUE))           ~ "Equipment & Gear",
      str_detect(Description, regex("Bagels|food", ignore_case = TRUE))                    ~ "Food & Refreshments",
      str_detect(Description, regex("Entry Fee|registration|Championship", ignore_case = TRUE)) ~ "Tournament & Entry Fees",
      str_detect(Description, regex("Certs|CPRO", ignore_case = TRUE))                      ~ "Training & Certification",
      str_detect(Description, regex("Indoor Courts", ignore_case = TRUE))                  ~ "Facility Rental",
      TRUE                                                                                 ~ "Miscellaneous"
    ),
    Spending = abs(Amount_Num)
  )

# Step 2: Summarize data for the pie chart.
spending_summary <- spending_categorized %>%
  group_by(Category) %>%
  summarise(TotalSpending = sum(Spending)) %>%
  mutate(Percentage = TotalSpending / sum(TotalSpending) * 100)

# Step 3: Create the pie chart.
ggplot(spending_summary, aes(x = "", y = TotalSpending, fill = Category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(round(Percentage), "%\n", "(", dollar(TotalSpending), ")")), 
            position = position_stack(vjust = 0.5), 
            color = "white", 
            size = 3.5) +
  labs(
    title = "Spending Distribution by Category",
    fill = "Spending Category"
  ) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```


```{r}
library(readxl); library(dplyr); library(ggplot2); library(scales); library(googlesheets4)

# TODO: swap back to read_sheet() when Sheet link is fixed
# Data = read_sheet(sheet_id)

Data = read_excel("data/TennisClubFinancialData.xlsx") %>%
  mutate(
    Date = as.Date(Date),
    YearMonth = format(Date, "%Y-%m")
  )

FinancialData = Data %>%
  mutate(
    Date = as.Date(Date),
    Year = format(Date, "%Y"),
    YearMonth = format(Date, "%Y-%m")
  )

ExpensesMonthly = FinancialData %>%
  filter(Flow == "Expense") %>%
  group_by(YearMonth, Category) %>%
  summarise(Spend = -sum(Amount, na.rm = TRUE), .groups="drop") %>%
  mutate(YearMonth = factor(YearMonth, levels = sort(unique(YearMonth))))

ColorPalette = c(
  "Travel"                 = "#F77F00",
  "Tournament Fees"        = "#00B4D8",
  "Facility Rentals"       = "#43AA8B",
  "Equipment & Supplies"   = "#577590",
  "Uniforms & Merchandise" = "#9D4EDD",
  "Miscellaneous/Admin"    = "#E5E5E5"
)
bg = "#861F41"; fg = "#FFFFFF"; grid = "#E0CFCF"

ggplot(ExpensesMonthly, aes(YearMonth, Spend, fill = Category)) +
  geom_col(width = 0.85) +
  labs(title = "Monthly Expense by Category",
       subtitle = "Aug 2024 – Present",
       x = "Month", y = "Spend (USD)", fill = "",
       caption = "Source: VT Club Tennis") +
  scale_y_continuous(labels = dollar_format(accuracy = 1),
                     expand = expansion(mult = c(0,.05))) +
  scale_fill_manual(values = ColorPalette) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background  = element_rect(fill = bg, color = bg),
    panel.background = element_rect(fill = bg, color = bg),
    panel.grid.major = element_line(color = grid, linewidth = 0.25),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 60, hjust = 1, color = fg),
    axis.text.y      = element_text(color = fg),
    axis.title.x     = element_text(color = fg, face = "bold"),
    axis.title.y     = element_text(color = fg, face = "bold"),
    plot.title       = element_text(hjust = 0.5, color = fg, face = "bold"),
    plot.subtitle    = element_text(hjust = 0.5, color = fg),
    plot.caption     = element_text(color = fg),
    legend.background= element_rect(fill = bg, color = bg),
    legend.key       = element_rect(fill = bg, color = bg),
    legend.text      = element_text(color = fg)
  )
```