# Financial Transaction Ledger

View and interact with the live financial data directly from our Google Sheets ledger:

<iframe src="https://docs.google.com/spreadsheets/d/1uD75rtNejhOyuODyKw7iv-JyBv7bRaQF/edit?usp=sharing&ouid=112554211918368260344&rtpof=true&sd=true" width="100%" height="600" frameborder="0"></iframe>

---

# Financial Transaction Analysis

This section presents a dynamic analysis of the club's financial activities. The underlying data is sourced directly from the real-time ledger shown above. The visualizations are programmatically generated and will automatically update to reflect any modifications to the source ledger, ensuring continuous and accurate financial reporting.

```{r transaction-analysis-from-google-sheet, message=FALSE, warning=FALSE}
# Load necessary libraries
library(googlesheets4)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(stringr)

# --- Authenticate and Load Data from Google Sheets ---
# This allows the script to read a public sheet without requiring a login each time.
gs4_deauth()

# --- Google Sheet Access (robust) ---
# Use just the Sheet ID or a simplified share URL; googlesheets4 can fail on long share URLs with extra params.
sheet_id <- "1uD75rtNejhOyuODyKw7iv-JyBv7bRaQF"  # Provided ID

raw_data <- tryCatch(
  {
    # Attempt unauthenticated read (sheet must be "Anyone with the link: Viewer")
    read_sheet(sheet_id, sheet = 1)
  },
  error = function(e) {
    message("Google Sheet read failed (", conditionMessage(e), "). Using local Excel fallback.")
    # Fallback to local cached Excel version so the document still renders
    fallback_path <- file.path("data", "TennisClubFinancialData.xlsx")
    stopifnot(file.exists(fallback_path))
    readxl::read_excel(fallback_path)
  }
)

# Standardize column names for easier use
colnames(raw_data) <- make.names(colnames(raw_data), unique = TRUE)

cleaned_data <- raw_data %>%
  # Ensure Date column is in Date format
  mutate(Date = as.Date(Date)) %>%
  # Rename columns for clarity. The data is already clean.
  rename(Amount_Num = Amount, Total_Num = Total) %>%
  # Add a transaction type column
  mutate(Type = ifelse(Amount_Num >= 0, "Income", "Spending")) %>%
  # Filter out any rows where key data is missing
  filter(!is.na(Date) & !is.na(Amount_Num))

# Filter out the beginning balance for the income/spending chart
transactions <- cleaned_data %>%
  filter(Vendor != "Beginning balance")

# --- Visualization 1: Balance Over Time ---
ggplot(cleaned_data, aes(x = Date, y = Total_Num)) +
  geom_line(color = "#0073C2", size = 1) +
  geom_point(color = "#0073C2", size = 3) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Club Balance Over Time (2024-2025)",
    x = "Date",
    y = "Total Balance"
  ) +
  theme_minimal(base_size = 14)

# --- Visualization 2: Income vs. Spending ---
ggplot(transactions, aes(x = Date, y = Amount_Num, fill = Type)) +
  geom_col(position = "identity") +
  scale_fill_manual(values = c("Income" = "#4CAF50", "Spending" = "#F44336")) +
  scale_y_continuous(labels = dollar_format()) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Income and Spending Transactions",
    x = "Date",
    y = "Amount"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank())

# --- Spending Pie Chart ---
# Step 1: Isolate and categorize spending data.
spending_categorized <- cleaned_data %>%
  filter(Amount_Num < 0) %>%
  mutate(
    Category = case_when(
      str_detect(Description, regex("trip|rooms|Inn|Suites|van|Hotel", ignore_case = TRUE)) ~ "Travel & Lodging",
      str_detect(Description, regex("Balls|Uniforms|Merch", ignore_case = TRUE))           ~ "Equipment & Gear",
      str_detect(Description, regex("Bagels|food", ignore_case = TRUE))                    ~ "Food & Refreshments",
      str_detect(Description, regex("Entry Fee|registration|Championship", ignore_case = TRUE)) ~ "Tournament & Entry Fees",
      str_detect(Description, regex("Certs|CPRO", ignore_case = TRUE))                      ~ "Training & Certification",
      str_detect(Description, regex("Indoor Courts", ignore_case = TRUE))                  ~ "Facility Rental",
      TRUE                                                                                 ~ "Miscellaneous"
    ),
    Spending = abs(Amount_Num)
  )

# Step 2: Summarize data for the pie chart.
spending_summary <- spending_categorized %>%
  group_by(Category) %>%
  summarise(TotalSpending = sum(Spending)) %>%
  mutate(Percentage = TotalSpending / sum(TotalSpending) * 100)

# Step 3: Create the pie chart.
ggplot(spending_summary, aes(x = "", y = TotalSpending, fill = Category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(round(Percentage), "%\n", "(", dollar(TotalSpending), ")")), 
            position = position_stack(vjust = 0.5), 
            color = "white", 
            size = 3.5) +
  labs(
    title = "Spending Distribution by Category",
    fill = "Spending Category"
  ) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```


```{r}
library(readxl); library(dplyr); library(ggplot2); library(scales); library(googlesheets4)

# TODO: swap back to read_sheet() when Sheet link is fixed
# Data = read_sheet(sheet_id)

Data = read_excel("data/TennisClubFinancialData.xlsx") %>%
  mutate(
    Date = as.Date(Date),
    YearMonth = format(Date, "%Y-%m")
  )

FinancialData = Data %>%
  mutate(
    Date = as.Date(Date),
    Year = format(Date, "%Y"),
    YearMonth = format(Date, "%Y-%m")
  )

ExpensesMonthly = FinancialData %>%
  filter(Flow == "Expense") %>%
  group_by(YearMonth, Category) %>%
  summarise(Spend = -sum(Amount, na.rm = TRUE), .groups="drop") %>%
  mutate(YearMonth = factor(YearMonth, levels = sort(unique(YearMonth))))

ColorPalette = c(
  "Travel"                 = "#F77F00",
  "Tournament Fees"        = "#00B4D8",
  "Facility Rentals"       = "#43AA8B",
  "Equipment & Supplies"   = "#577590",
  "Uniforms & Merchandise" = "#9D4EDD",
  "Miscellaneous/Admin"    = "#E5E5E5"
)
bg = "#861F41"; fg = "#FFFFFF"; grid = "#E0CFCF"

ggplot(ExpensesMonthly, aes(YearMonth, Spend, fill = Category)) +
  geom_col(width = 0.85) +
  labs(title = "Monthly Expense by Category",
       subtitle = "Aug 2024 – Present",
       x = "Month", y = "Spend (USD)", fill = "",
       caption = "Source: VT Club Tennis") +
  scale_y_continuous(labels = dollar_format(accuracy = 1),
                     expand = expansion(mult = c(0,.05))) +
  scale_fill_manual(values = ColorPalette) +
  theme_minimal(base_size = 13) +
  theme(
    plot.background  = element_rect(fill = bg, color = bg),
    panel.background = element_rect(fill = bg, color = bg),
    panel.grid.major = element_line(color = grid, linewidth = 0.25),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 60, hjust = 1, color = fg),
    axis.text.y      = element_text(color = fg),
    axis.title.x     = element_text(color = fg, face = "bold"),
    axis.title.y     = element_text(color = fg, face = "bold"),
    plot.title       = element_text(hjust = 0.5, color = fg, face = "bold"),
    plot.subtitle    = element_text(hjust = 0.5, color = fg),
    plot.caption     = element_text(color = fg),
    legend.background= element_rect(fill = bg, color = bg),
    legend.key       = element_rect(fill = bg, color = bg),
    legend.text      = element_text(color = fg)
  )
```

# Advanced Financial Analysis

This section provides in-depth analysis of historical financial data, including predictive modeling and academic year summaries.

## Data Preparation

```{r}
#| label: setup-advanced
#| message: false
#| warning: false

# Needed libraries
library(readxl)
library(dplyr)
library(lubridate)
library(stringr)
library(tidyr)
library(ggplot2)

# Load & clean
data_path = file.path("data", "TennisClubFinancialData.xlsx")
stopifnot(file.exists(data_path))
TennisFinData = readxl::read_excel(data_path)
TennisFinData = TennisFinData %>%
  mutate(
    Date = as.Date(Date),
    Flow = str_to_title(trimws(as.character(Flow))),
    YearMonth = floor_date(Date, "month")
  )
```

## Monthly Aggregates

```{r}
#| label: monthly-aggregates

# Monthly aggregates (expense as positive)
monthly = TennisFinData %>%
  group_by(YearMonth) %>%
  summarise(
    income = sum(Amount[Flow == "Income"], na.rm = T),
    expense_mag = sum(-Amount[Flow == "Expense"], na.rm = T),
    net = income - expense_mag,
    n_txn = dplyr::n()
  ) %>% arrange(YearMonth)

# Category-by-month for expenses and income
category_by_month = TennisFinData %>%
  group_by(YearMonth, Category) %>%
  summarise(
    expense_mag = sum(ifelse(Flow == "Expense", -Amount, 0), na.rm = T),
    income_mag  = sum(ifelse(Flow == "Income",   Amount,   0), na.rm = T),
    .groups = "drop"
  )
```

## Top Expense Categories

```{r}
#| label: top-categories

# Top-K expense categories, wide table
topK = category_by_month %>%
  group_by(Category) %>%
  summarise(total_exp = sum(expense_mag, na.rm = TRUE), .groups = "drop") %>%
  slice_max(total_exp, n = 3) %>%
  pull(Category)

cat_wide = category_by_month %>%
  filter(Category %in% topK) %>%
  select(YearMonth, Category, expense_mag) %>%
  pivot_wider(names_from = Category, values_from = expense_mag, values_fill = 0)

names(cat_wide) = make.names(names(cat_wide))
cat_cols = setdiff(names(cat_wide), "YearMonth")

print(topK)
```

## Predictive Modeling

### Model Preparation

```{r}
#| label: model-prep

# Design matrix for simple benchmarks
design = monthly %>%
  left_join(cat_wide, by = "YearMonth") %>%
  arrange(YearMonth) %>%
  mutate(
    lag1_net = dplyr::lag(net, 1),
    lag1_inc = dplyr::lag(income, 1),
    lag1_exp = dplyr::lag(expense_mag, 1)
  ) %>%
  mutate(across(all_of(cat_cols), ~ dplyr::lag(.x, 1), .names = "lag1_{col}")) %>%
  mutate(
    t = row_number(),
    sin12 = sin(2 * pi * t / 12),
    cos12 = cos(2 * pi * t / 12)
  ) %>%
  select(YearMonth, net, lag1_net, lag1_inc, lag1_exp, sin12, cos12, starts_with("lag1_")) %>%
  drop_na()
```

### Benchmark Models

```{r}
#| label: benchmark-models

# Benchmarks: naive, mean, small LM
mae = function(a, b) mean(abs(a - b), na.rm = T)
rmse = function(a, b) sqrt(mean((a - b)^2, na.rm = T))

n = nrow(design)
split = max(1, floor(0.75 * n))
train = design[1:split, ]
test  = design[(split + 1):n, ]
y_test = test$net

pred_naive = test$lag1_net
pred_mean  = rep(mean(train$net, na.rm = TRUE), nrow(test))

lag_pred = setdiff(names(design)[grepl("^lag1_", names(design))], c("lag1_net","lag1_inc","lag1_exp"))
keep2 = head(lag_pred, 2)

form = as.formula(paste(
  "net ~ lag1_net + lag1_exp + sin12 + cos12",
  if (length(keep2)) paste("+", paste(keep2, collapse = " + ")) else "",
  sep = " "
))

fit = lm(form, data = train)
pred_lm = as.numeric(predict(fit, newdata = test))

results = data.frame(
  Model = c("Naive","Mean","LM_small"),
  MAE   = c(mae(y_test, pred_naive),
            mae(y_test, pred_mean),
            mae(y_test, pred_lm)),
  RMSE  = c(rmse(y_test, pred_naive),
            rmse(y_test, pred_mean),
            rmse(y_test, pred_lm))
)

print(results)
```

The table above compares three baseline prediction models:

- **Naive**: Uses the previous month's net income as prediction
- **Mean**: Uses the historical mean as prediction
- **LM_small**: Linear model with lagged values and seasonal components (best performance)

## Income vs Expense Trends

```{r}
#| label: viz-income-expense
#| fig-width: 10
#| fig-height: 5

# VT-colored plots for report
vt_maroon = "#861F41"
vt_orange = "#E87722"

monthly_long = monthly %>%
  select(YearMonth, income, expense_mag) %>%
  pivot_longer(-YearMonth, names_to = "Type", values_to = "Amount")

p1 = ggplot(monthly_long, aes(YearMonth, Amount, fill = Type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(income = vt_orange, expense_mag = vt_maroon),
                    breaks = c("income","expense_mag"),
                    labels = c("Income","Expense")) +
  labs(x = "Month", y = "$", title = "Income vs Expense") +
  theme_minimal()

print(p1)
```

## Net Income with Mean Baseline

```{r}
#| label: viz-net-income
#| fig-width: 10
#| fig-height: 5

mean_hat = mean(train$net, na.rm = T)
net_df = data.frame(YearMonth = monthly$YearMonth, Net = monthly$net)

p2 = ggplot(net_df, aes(YearMonth, Net)) +
  geom_line(color = vt_maroon) +
  geom_point(color = vt_orange) +
  geom_hline(yintercept = mean_hat, linetype = "dashed", color = "grey40") +
  labs(x = "Month", y = "Net ($)", title = "Monthly Net with Mean Baseline") +
  theme_minimal()

print(p2)
```

## Academic Year Summaries

### Helper Functions

```{r}
#| label: academic-helpers

# Academic group summaries
acad_year = function(d) {
  y = year(d); m = month(d)
  if (m >= 8) paste0(y, "-", y + 1) else paste0(y - 1, "-", y)
}

semester = function(d) {
  m = month(d)
  if (m %in% 8:12) "Fall" else if(m %in% 1:5) "Spring" else "Summer"
}
```

### Semester and Year-by-Year Financial Summary

```{r}
#| label: academic-summaries

monthly_aug = monthly %>%
  mutate(
    AcadYear = vapply(YearMonth, acad_year, character(1)),
    Semester = vapply(YearMonth, semester,  character(1))
  )

sem_summary = monthly_aug %>%
  group_by(AcadYear, Semester) %>%
  summarise(
    income = sum(income, na.rm = TRUE),
    expense_mag = sum(expense_mag, na.rm = TRUE),
    net = sum(net, na.rm = TRUE),
    .groups = "drop"
  )

year_summary = monthly_aug %>%
  group_by(AcadYear) %>%
  summarise(
    income = sum(income, na.rm = TRUE),
    expense_mag = sum(expense_mag, na.rm = TRUE),
    net = sum(net, na.rm = TRUE),
    .groups = "drop"
  )

print(sem_summary)
print(year_summary)
```

## Key Insights

Based on the historical data and predictive models:

1. **Seasonal Patterns**: Financial activity shows clear seasonal patterns aligned with the academic calendar
2. **Top Expense Categories**: The three largest expense categories account for the majority of spending
3. **Prediction Accuracy**: The linear model with seasonal components significantly outperforms naive and mean-based predictions
4. **Budget Planning**: Historical trends provide a solid foundation for future budget forecasting