---
title: "Project Log"
subtitle: "Meeting Notes & Progress Updates"
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(knitr)

# VT Colors
vt_maroon <- "#861F41"
vt_orange <- "#E87722"
```

This page documents our ongoing collaboration with VT Club Tennis, tracking meetings, decisions, and project milestones.

---

## Project Timeline Visualization

```{r timeline-viz, echo=FALSE, fig.width=10, fig.height=6}
# Create timeline data
timeline_data <- data.frame(
  phase = c("Initial Meeting", "Requirements", "Development", "Testing", "Delivery"),
  start_date = as.Date(c("2025-09-01", "2025-09-15", "2025-10-01", "2025-11-01", "2025-11-25")),
  end_date = as.Date(c("2025-09-14", "2025-09-30", "2025-10-31", "2025-11-24", "2025-12-10")),
  status = c("Completed", "Completed", "In Progress", "Upcoming", "Upcoming"),
  y_pos = c(1, 2, 3, 4, 5)
)

# Create the timeline visualization
ggplot(timeline_data, aes(x = start_date, y = y_pos)) +
  geom_segment(aes(x = start_date, xend = end_date, y = y_pos, yend = y_pos, color = status),
               size = 8, lineend = "round") +
  geom_point(aes(x = start_date, color = status), size = 5) +
  geom_point(aes(x = end_date, color = status), size = 5) +
  geom_text(aes(label = phase, x = start_date + (end_date - start_date)/2), 
            vjust = -1.5, size = 4, fontface = "bold") +
  scale_color_manual(values = c("Completed" = vt_maroon, 
                                 "In Progress" = vt_orange,
                                 "Upcoming" = "gray60")) +
  scale_y_continuous(limits = c(0.5, 5.5)) +
  labs(title = "Project Timeline",
       subtitle = "Fall 2025 Semester",
       x = "Date",
       y = "",
       color = "Status") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.title = element_text(size = 16, face = "bold", color = vt_maroon),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    legend.position = "bottom"
  )
```

---

## Progress Overview

```{r progress-viz, echo=FALSE, fig.width=10, fig.height=5}
# Create deliverables progress data
deliverables <- data.frame(
  item = c("Financial Analysis", "Budget Tool", "Elo Ranking", 
           "Performance Analytics", "Final Presentation"),
  progress = c(85, 90, 75, 60, 20),
  category = c("Financial", "Financial", "Performance", "Performance", "General")
)

deliverables$item <- factor(deliverables$item, levels = rev(deliverables$item))

ggplot(deliverables, aes(x = progress, y = item, fill = category)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(progress, "%")), hjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Financial" = vt_maroon, 
                                "Performance" = vt_orange,
                                "General" = "gray60")) +
  scale_x_continuous(limits = c(0, 110), expand = c(0, 0)) +
  labs(title = "Deliverables Progress",
       subtitle = "Completion Status",
       x = "Progress (%)",
       y = "",
       fill = "Category") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", color = vt_maroon),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    legend.position = "bottom",
    panel.grid.major.y = element_blank()
  )
```

---

## Meeting 4 - [Date]

### Attendees
- 

### Topics Discussed
- 

### Decisions Made
- 

### Action Items
- [ ] 

### Next Steps
- 

---

## Meeting 3 - [Date]

### Attendees
- 

### Topics Discussed
- 

### Decisions Made
- 

### Action Items
- [ ] 

### Next Steps
- 

---

## Meeting 2 - [Date]

### Attendees
- 

### Topics Discussed
- 

### Decisions Made
- 

### Action Items
- [ ] 

### Next Steps
- 

---

## Meeting 1 - [Date]

### Attendees
- 

### Topics Discussed
- 

### Decisions Made
- 

### Action Items
- [ ] 

### Next Steps
- 

---

## Key Milestones Summary

```{r milestones-table, echo=FALSE}
milestones <- data.frame(
  Date = c("Sept 1", "Sept 15", "Oct 1", "Nov 1", "Nov 25", "Dec 10"),
  Milestone = c(
    "Initial Client Meeting",
    "Requirements Gathering Complete",
    "Development Phase Begins",
    "Testing Phase Begins",
    "Client Review & Feedback",
    "Final Delivery"
  ),
  Status = c("âœ… Completed", "âœ… Completed", "ðŸ”„ In Progress", 
             "â³ Upcoming", "â³ Upcoming", "â³ Upcoming")
)

kable(milestones, align = c('l', 'l', 'c'))
```

---

*Last Updated: November 18, 2025*
